---
title: "Exploratory Analysis & Normalization of COPD RNA-seq Data (GSE57148)"
author: "Based on Kim et al., 2015 – Int J Genomics"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 9, fig.height = 6,
                      fig.path = "fig/copd-", warning = FALSE, message = FALSE)
```

# Introduction

We perform exploratory gene expression analysis and
normalization on the **GSE57148** dataset from Kim et al. (2015),
*"Comprehensive Analysis of Transcriptome Sequencing Data in the Lung Tissues
of COPD Subjects"*.

The dataset contains RNA-seq raw counts from lung tissue resections of:

- **91 control subjects** with normal spirometry
- **98 COPD subjects**

All subjects are male, sequenced on the Illumina HiSeq 2000 platform.

### Paper normalization pipeline

The paper's Methods section states:

> *"Expression levels were extracted as a **FPKM** value for each gene of each
> sample using Cufflinks software. Genes with FPKM values of 0 across all
> samples were excluded. Filtered data were subject to **upper quantile
> normalization**."*

We reproduce this exact two-step pipeline from the raw counts:

1. **FPKM** — within-sample normalization correcting for gene length and
   sequencing depth
2. **Upper Quartile** — between-sample global normalization correcting for
   residual depth differences across samples

We also run TMM, RLE, CPM, and Full Quantile to compare against the paper's
choice in the evaluation plots.

**Files required to knit this document:**

- `GSE57148_raw_counts_GRCh38_p13_NCBI.tsv`
- `GSE57148_series_matrix.txt`
- `Human_GRCh38_p13_annot.tsv`

---

## R packages installation

```{r, eval=FALSE}
install.packages(c("dplyr", "ggplot2", "corrplot", "ggfortify"))
if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("limma", "EDASeq", "edgeR", "SummarizedExperiment"))
```

## Load packages

```{r, message=FALSE}
library(limma)
library(EDASeq)
library(edgeR)
library(SummarizedExperiment)
library(dplyr)
library(ggplot2)
library(corrplot)
library(ggfortify)
```

---

# Data Loading

## Load sample metadata from the series matrix

The series matrix file confirms the true COPD/Control group labels directly
from the GEO submission.

```{r}
sm_lines <- readLines("GSE57148_series_matrix.txt")

acc_line     <- sm_lines[grep("^!Sample_geo_accession", sm_lines)]
gsm_ids      <- gsub('"', '', strsplit(acc_line, "\t")[[1]][-1])

title_line   <- sm_lines[grep("^!Sample_title", sm_lines)]
titles       <- gsub('"', '', strsplit(title_line, "\t")[[1]][-1])

disease_line <- sm_lines[grep("^!Sample_characteristics_ch1.*disease state",
                               sm_lines)[1]]
disease_vals <- gsub("disease state: ", "",
                     gsub('"', '', strsplit(disease_line, "\t")[[1]][-1]))

sample_meta <- data.frame(
  gsm     = gsm_ids,
  title   = titles,
  disease = disease_vals,
  stringsAsFactors = FALSE
)
sample_meta$group <- factor(
  ifelse(sample_meta$disease == "COPD", "COPD", "Control"),
  levels = c("Control", "COPD")
)

cat("Group counts confirmed from series matrix:\n")
print(table(sample_meta$group))
```

## Load gene annotation table

The annotation file provides two essential columns:

- `GeneType` — to filter out non-coding genes (as the paper did)
- `Length` — gene length in bp, required for FPKM computation

```{r}
annot <- read.table(
  "Human_GRCh38_p13_annot.tsv",
  header = TRUE, sep = "\t", quote = "", fill = TRUE, comment.char = ""
)
cat("Annotation dimensions:", dim(annot), "\n")
cat("Gene types available:\n")
print(sort(table(annot$GeneType), decreasing = TRUE))
```

## Load raw count matrix

```{r}
raw_data <- read.table(
  "GSE57148_raw_counts_GRCh38_p13_NCBI.tsv",
  header = TRUE, sep = "\t", row.names = 1
)
cat("Raw count matrix dimensions (genes x samples):", dim(raw_data), "\n")
raw_data[1:4, 1:4]
```

## Align samples

```{r}
common_samples <- intersect(sample_meta$gsm, colnames(raw_data))
raw_data       <- raw_data[, common_samples]
sample_meta    <- sample_meta[match(common_samples, sample_meta$gsm), ]
stopifnot(all(colnames(raw_data) == sample_meta$gsm))

col_group <- ifelse(sample_meta$group == "COPD", "#E05C5C", "#5B8DB8")
n_control <- sum(sample_meta$group == "Control")
n_copd    <- sum(sample_meta$group == "COPD")
cat("Samples aligned:", ncol(raw_data), "\n")
```

---

# Build SummarizedExperiment

```{r}
se <- SummarizedExperiment(
  assays  = list(counts = as.matrix(raw_data)),
  colData = DataFrame(sample_meta)
)
se
```

---

# Pre-filtering

## Remove all-zero genes

```{r}
zero_genes <- rowSums(assay(se, "counts")) == 0
cat("All-zero genes removed:", sum(zero_genes), "\n")
se <- se[!zero_genes, ]
```

## Remove non-coding genes

Following the paper, only **protein-coding** genes are retained.

```{r}
gene_ids  <- as.integer(rownames(se))
annot_sub <- annot[match(gene_ids, annot$GeneID), ]
is_coding <- !is.na(annot_sub$GeneType) & annot_sub$GeneType == "protein-coding"
cat("Protein-coding genes retained:", sum(is_coding), "out of", nrow(se), "\n")

se        <- se[is_coding, ]
annot_sub <- annot_sub[is_coding, ]        # keep annotation aligned to matrix
rowData(se)$symbol <- annot_sub$Symbol
rowData(se)$length <- annot_sub$Length     # gene length in bp — needed for FPKM

raw_count <- assay(se, "counts")
group     <- colData(se)$group
cat("Final matrix:", nrow(raw_count), "genes x", ncol(raw_count), "samples\n")
```

## Additional low-count filtering

Genes with a row sum below 10 are removed to reduce noise from very lowly
expressed genes before normalization.

```{r}
keep      <- rowSums(raw_count) >= 10
cat("Low-count genes removed:", sum(!keep), "| Remaining:", sum(keep), "\n")

se        <- se[keep, ]
annot_sub <- annot_sub[keep, ]
raw_count <- assay(se, "counts")
group     <- colData(se)$group
```

---

# Exploratory Analysis — Before Normalization

Before normalizing, we visualize the data to identify samples with unusually
low depth, potential batch effects, or outliers. These plots also serve as a
baseline to compare against post-normalization results.

## Counts distribution

### Raw counts histogram

Raw counts are highly right-skewed — most genes have very few counts and a
handful dominate with very high counts.

```{r}
hist(raw_count[, 1],
     breaks = 100, col = "#5B8DB8", border = "white",
     main = paste0("Raw counts – ", colnames(raw_count)[1]),
     xlab = "Raw counts",
     xlim = c(0, quantile(raw_count[, 1], 0.99)))
```

### Pseudo-counts (log1p) histogram

`log(count + 1)` compresses the dynamic range, handles zeros, and gives a
more symmetric, interpretable distribution.

```{r}
hist(log1p(raw_count[, 1]),
     breaks = 40, col = "#5B8DB8", border = "white",
     main = paste0("Pseudo-counts (log1p) – ", colnames(raw_count)[1]),
     xlab = expression(log[2](count + 1)))
```

---

## Boxplot

```{r, fig.width=18, fig.height=6}
idx <- 1:40
boxplot(log1p(raw_count[, idx]),
        col = col_group[idx],
        names = substr(colnames(raw_count)[idx], 4, 10),
        las = 2, cex.axis = 0.65,
        ylab = expression(log[1+p](count)),
        main = "Boxplot of pseudo-counts (first 40 samples)",
        outline = FALSE)
legend("topright", legend = c("Control", "COPD"),
       fill = c("#5B8DB8", "#E05C5C"), bty = "n")
abline(h = median(log1p(raw_count)), lty = 2, col = "grey50")
```

```{r, fig.width=22, fig.height=6}
boxplot(log1p(raw_count),
        col = col_group, names = NA, outline = FALSE,
        ylab = expression(log[1+p](count)),
        main = "Boxplot of pseudo-counts – all 189 samples")
legend("topright", legend = c("Control", "COPD"),
       fill = c("#5B8DB8", "#E05C5C"), bty = "n")
abline(v = n_control + 0.5, lty = 2, lwd = 1.5)
text(n_control / 2,          max(log1p(raw_count)) * 0.97, "Control")
text(n_control + n_copd / 2, max(log1p(raw_count)) * 0.97, "COPD")
```

---

## Library size (sequencing depth)

```{r, fig.width=14, fig.height=5}
lib_size <- colSums(raw_count) / 1e6
barplot(lib_size, col = col_group, names.arg = NA, border = NA,
        ylab = "Library size (M reads)", main = "Library sizes per sample")
legend("topright", legend = c("Control", "COPD"),
       fill = c("#5B8DB8", "#E05C5C"), bty = "n")
abline(h = mean(lib_size), lty = 2, col = "grey40")
abline(v = n_control + 0.5, lty = 2, lwd = 1.5)
cat("Mean library size:", round(mean(lib_size), 2), "M reads\n")
```

---

## MD / MA plot

A well-behaved sample scatters symmetrically around M = 0 relative to a
pseudo-reference built as the row-wise median across all samples.

```{r, fig.width=12, fig.height=5}
par(mfrow = c(1, 2))
limma::plotMA(log1p(raw_count), array = 1,
              main = paste0("MA – Control\n", colnames(raw_count)[1]),
              xlab = "A (mean log expression)", ylab = "M (log fold-change)")
abline(h = 0, col = "blue", lty = 2)
limma::plotMA(log1p(raw_count), array = n_control + 1,
              main = paste0("MA – COPD\n", colnames(raw_count)[n_control + 1]),
              xlab = "A (mean log expression)", ylab = "M (log fold-change)")
abline(h = 0, col = "red", lty = 2)
par(mfrow = c(1, 1))
```

---

## RLE plot — before normalization

$$z_{ij} = \log\!\left(\frac{x_{ij}}{\text{Med}(\tilde{x}_i)}\right)$$

Each sample's distribution should be centred on zero with small, uniform
spread. Deviations indicate technical bias that normalization must correct.

```{r, fig.width=20, fig.height=6}
EDASeq::plotRLE(raw_count,
                col = col_group, outline = FALSE, las = 2, cex.axis = 0.5,
                ylab = "RLE", main = "RLE – raw counts (before normalization)")
abline(h = 0, lty = 2, col = "grey40")
abline(v = n_control + 0.5, lty = 2, lwd = 1.5)
legend("topright", legend = c("Control", "COPD"),
       fill = c("#5B8DB8", "#E05C5C"), bty = "n")
```

---

## Mean–variance relationship

RNA-seq count data show **overdispersion**: variance grows faster than the
mean (quadratic rather than linear). This motivates negative-binomial models
(edgeR / DESeq2) for DE analysis.

```{r, fig.width=7, fig.height=6}
smoothScatter(log10(rowMeans(raw_count) + 1),
              log10(apply(raw_count, 1, var) + 1),
              xlab = expression(log[10](mean + 1)),
              ylab = expression(log[10](variance + 1)),
              main = "Mean–Variance relationship (raw counts)")
abline(0, 1, col = "purple", lwd = 2)
abline(0, 2, col = "orange", lwd = 2, lty = 2)
legend("topleft", legend = c("Linear (Poisson)", "Quadratic (overdispersion)"),
       col = c("purple", "orange"), lwd = 2, lty = c(1, 2), bty = "n")
```

---

## PCA — before normalization

```{r}
pca_raw <- prcomp(t(log1p(raw_count)), scale. = FALSE)
var_raw <- summary(pca_raw)$importance[2, 1:2] * 100

screeplot(pca_raw, npcs = 20, type = "lines",
          main = "Scree plot – raw counts")
```

```{r}
pca_df <- data.frame(PC1 = pca_raw$x[,1], PC2 = pca_raw$x[,2], group = group)
ggplot(pca_df, aes(PC1, PC2, colour = group)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_manual(values = c("Control" = "#5B8DB8", "COPD" = "#E05C5C")) +
  stat_ellipse(level = 0.95, linetype = 2) +
  labs(title = "PCA – raw counts (before normalization)",
       x = sprintf("PC1 (%.1f%%)", var_raw[1]),
       y = sprintf("PC2 (%.1f%%)", var_raw[2]), colour = "Group") +
  theme_bw(base_size = 13)
```

---

# Normalization

Every measurement technique is affected by systematic errors — biases that
cannot be reduced by increasing sample size and must be eliminated by
normalization. The goal is to make differences between samples reflect only
biological variation.

The main sources of bias in RNA-seq are: **sequencing depth**, **gene length**,
**GC content**, and **library preparation differences**.

## Paper pipeline: FPKM + Upper Quartile

### Step 1 — FPKM (within-sample)

**FPKM** (Fragments Per Kilobase of exon per Million mapped reads) is a
non-global within-sample normalization that corrects simultaneously for:

- **Gene length** — longer genes accumulate more reads by chance regardless
  of expression level
- **Sequencing depth** — samples with more total reads get proportionally
  higher counts in every gene

$$\text{FPKM}_{ij} = \frac{x_{ij}}{\text{Length}_i[\text{kb}] \times N_j[\text{M reads}]}$$

where $x_{ij}$ is the raw count for gene $i$ in sample $j$,
$\text{Length}_i$ is the gene length in kilobases from the annotation, and
$N_j$ is the total mapped reads for sample $j$ in millions.

```{r}
# Gene lengths in kilobases
gene_length_kb <- rowData(se)$length / 1000

# Library sizes in millions of mapped reads
lib_size_M <- colSums(raw_count) / 1e6

# Compute FPKM: divide each row by its gene length, each column by lib size
fpkm <- sweep(raw_count, 1, gene_length_kb, "/")   # ÷ gene length (kb)
fpkm <- sweep(fpkm,      2, lib_size_M,    "/")    # ÷ library size (M reads)

cat("FPKM matrix dimensions:", dim(fpkm), "\n")
cat("FPKM value range (sample 1):", round(range(fpkm[, 1]), 4), "\n")
```

### Step 2 — Upper Quartile normalization (between-sample)

After FPKM, the paper applied **upper quartile normalization** to correct
residual between-sample differences in sequencing depth. This is a global
between-sample method: a single scaling factor is estimated per sample and
applied to all genes in that sample.

The scaling factor is the **75th percentile** (upper quartile) of the
non-zero FPKM values in each sample. Dividing by the upper quartile (rather
than the total sum / median) is more robust because it is less sensitive to
the small fraction of very highly expressed genes that dominate the total count.

$$\text{FPKM-UQ}_{ij} = \frac{\text{FPKM}_{ij}}{Q_{75,j}} \times 10^6$$

```{r}
# Compute the 75th percentile of non-zero FPKM values per sample
uq_factors <- apply(fpkm, 2, function(x) quantile(x[x > 0], 0.75))

cat("Upper quartile factors (first 6 samples):\n")
print(round(uq_factors[1:6], 6))

# Scale each sample and multiply by 1e6 to keep values in a readable range
fpkm_uq <- sweep(fpkm, 2, uq_factors / 1e6, "/")

cat("\nFPKM + Upper Quartile matrix dimensions:", dim(fpkm_uq), "\n")
cat("This is the paper's final normalized matrix.\n")
```

---

## Additional normalizations for comparison

We also compute the standard between-sample methods (applied
directly to raw counts) to compare against the paper's approach.

```{r}
# Total Count (CPM) — simplest: divides only by library size
tc <- cpm(raw_count)

# Upper Quartile on raw counts (edgeR)
lbs_uq <- calcNormFactors(raw_count, method = "upperquartile")
uq_raw <- cpm(raw_count, lib.size = lbs_uq * colSums(raw_count))

# TMM — Trimmed Mean of M-values (edgeR default, robust global method)
lbs_tmm <- calcNormFactors(raw_count, method = "TMM")
tmm     <- cpm(raw_count, lib.size = lbs_tmm * colSums(raw_count))

# RLE — Geometric Mean (similar to DESeq2 median-of-ratios)
lbs_rle <- calcNormFactors(raw_count, method = "RLE")
rle     <- cpm(raw_count, lib.size = lbs_rle * colSums(raw_count))

# Full Quantile (EDASeq) — forces identical distributions across all samples
set_fq <- newSeqExpressionSet(
  raw_count,
  phenoData = AnnotatedDataFrame(
    data.frame(conditions = factor(group), row.names = colnames(raw_count))
  )
)
set_fq <- betweenLaneNormalization(set_fq, which = "full", offset = TRUE)
full_q <- set_fq@assayData$normalizedCounts

cat("All normalization matrices computed.\n")
```

# Evaluation Plots

## Boxplot comparison

```{r, fig.width=22, fig.height=16}
par(mfrow = c(3, 3))

boxplot(log1p(raw_count), col = col_group, names = NA, outline = FALSE,
        ylab = "log1p(count)",   main = "Raw counts")
boxplot(log1p(fpkm),      col = col_group, names = NA, outline = FALSE,
        ylab = "log1p(FPKM)",   main = "FPKM (within-sample)")
boxplot(log1p(fpkm_uq),   col = col_group, names = NA, outline = FALSE,
        ylab = "log1p(FPKM-UQ)", main = "FPKM + Upper Quartile  ← PAPER")
boxplot(log2(tc + 1),     col = col_group, names = NA, outline = FALSE,
        ylab = "log2(count+1)", main = "Total Count (CPM)")
boxplot(log2(uq_raw + 1), col = col_group, names = NA, outline = FALSE,
        ylab = "log2(count+1)", main = "Upper Quartile (raw counts)")
boxplot(log2(tmm + 1),    col = col_group, names = NA, outline = FALSE,
        ylab = "log2(count+1)", main = "TMM")
boxplot(log2(rle + 1),    col = col_group, names = NA, outline = FALSE,
        ylab = "log2(count+1)", main = "RLE")
boxplot(log2(full_q + 1), col = col_group, names = NA, outline = FALSE,
        ylab = "log2(count+1)", main = "Full Quantile")

par(mfrow = c(1, 1))
```

## RLE plot comparison

The RLE plot is a more sensitive tool than the boxplot. After good normalization all medians should be near zero with small, uniform spread. Comparing methods side-by-side makes it clear which normalizations best remove technical bias.

```{r, fig.width=20, fig.height=6}
EDASeq::plotRLE(raw_count, col = col_group, outline = FALSE,
                las = 2, cex.axis = 0.4, ylab = "RLE",
                main = "RLE – Raw counts (before normalization)")
abline(h = 0, lty = 2, col = "grey40"); abline(v = n_control + 0.5, lty = 2)
```

```{r, fig.width=20, fig.height=6}
EDASeq::plotRLE(fpkm, col = col_group, outline = FALSE,
                las = 2, cex.axis = 0.4, ylab = "RLE",
                main = "RLE – FPKM (within-sample only)")
abline(h = 0, lty = 2, col = "grey40"); abline(v = n_control + 0.5, lty = 2)
```

```{r, fig.width=20, fig.height=6}
EDASeq::plotRLE(fpkm_uq, col = col_group, outline = FALSE,
                las = 2, cex.axis = 0.4, ylab = "RLE",
                main = "RLE – FPKM + Upper Quartile  ← PAPER METHOD")
abline(h = 0, lty = 2, col = "grey40"); abline(v = n_control + 0.5, lty = 2)
legend("topright", legend = c("Control", "COPD"),
       fill = c("#5B8DB8", "#E05C5C"), bty = "n")
```

```{r, fig.width=20, fig.height=6}
EDASeq::plotRLE(tmm, col = col_group, outline = FALSE,
                las = 2, cex.axis = 0.4, ylab = "RLE",
                main = "RLE – TMM")
abline(h = 0, lty = 2, col = "grey40"); abline(v = n_control + 0.5, lty = 2)
```

```{r, fig.width=20, fig.height=6}
EDASeq::plotRLE(rle, col = col_group, outline = FALSE,
                las = 2, cex.axis = 0.4, ylab = "RLE",
                main = "RLE – RLE (Geometric Mean)")
abline(h = 0, lty = 2, col = "grey40"); abline(v = n_control + 0.5, lty = 2)
```

```{r, fig.width=20, fig.height=6}
EDASeq::plotRLE(full_q, col = col_group, outline = FALSE,
                las = 2, cex.axis = 0.4, ylab = "RLE",
                main = "RLE – Full Quantile")
abline(h = 0, lty = 2, col = "grey40"); abline(v = n_control + 0.5, lty = 2)
```

---

## PCA comparison

After normalization, PCA shows whether the known biological grouping
(COPD vs Control) is more clearly resolved. The normalization that best
separates the groups while keeping replicates together is preferred.

```{r}
pca_plot <- function(mat, title, grp = group) {
  p  <- prcomp(t(log1p(mat)), scale. = FALSE)
  ve <- summary(p)$importance[2, 1:2] * 100
  df <- data.frame(PC1 = p$x[,1], PC2 = p$x[,2], group = grp)
  ggplot(df, aes(PC1, PC2, colour = group)) +
    geom_point(size = 2, alpha = 0.8) +
    scale_colour_manual(values = c("Control" = "#5B8DB8", "COPD" = "#E05C5C")) +
    stat_ellipse(level = 0.95, linetype = 2) +
    labs(title = title,
         x = sprintf("PC1 (%.1f%%)", ve[1]),
         y = sprintf("PC2 (%.1f%%)", ve[2]),
         colour = "Group") +
    theme_bw(base_size = 10)
}

print(pca_plot(raw_count, "Raw counts"))
print(pca_plot(fpkm,      "FPKM (within-sample)"))
print(pca_plot(fpkm_uq,   "FPKM + Upper Quartile  ← PAPER"))
print(pca_plot(uq_raw,    "Upper Quartile (raw counts)"))
print(pca_plot(tmm,       "TMM"))
print(pca_plot(rle,       "RLE"))
print(pca_plot(full_q,    "Full Quantile"))
```

---

## Correlation plot — paper normalization (FPKM + UQ)

High within-group correlations confirm that the normalization has not distorted
the data. Low-correlation samples relative to their peers may be outliers.

```{r, fig.width=10, fig.height=9}
vars    <- apply(log1p(fpkm_uq), 1, var)
top5k   <- order(vars, decreasing = TRUE)[1:5000]
cor_mat <- cor(log1p(fpkm_uq[top5k, ]), method = "pearson")

ctrl_idx <- which(group == "Control")
copd_idx <- which(group == "COPD")

cat("Mean within-Control correlation (FPKM-UQ):",
    round(mean(cor_mat[ctrl_idx, ctrl_idx][
      upper.tri(cor_mat[ctrl_idx, ctrl_idx])]), 4), "\n")
cat("Mean within-COPD correlation (FPKM-UQ):   ",
    round(mean(cor_mat[copd_idx, copd_idx][
      upper.tri(cor_mat[copd_idx, copd_idx])]), 4), "\n")

col_pal <- colorRampPalette(c("#2166AC", "#F7F7F7", "#D6604D"))(200)
corrplot(cor_mat,
         method = "color", col = col_pal, tl.pos = "n",
         title = "Sample–sample correlation – FPKM + Upper Quartile (paper method)",
         mar = c(0, 0, 2, 0), addgrid.col = NA)
```


# Save data for next tutorial

```{r}
save(list = c("se", "raw_count", "fpkm", "fpkm_uq",
              "tmm", "rle", "tc", "uq_raw", "full_q",
              "group", "sample_meta", "annot_sub"),
     file = "normalization.RData")

cat("Saved normalization.RData\n")
cat("  fpkm_uq  = FPKM + Upper Quartile (paper method)\n")
cat("  tmm      = TMM on raw counts (recommended for DE)\n")
```

